Reducing Acceptance Marks in Emerson-Lei Automata by QBF Solving

### 2023-06-01 paper/main.tex
progress mainly in Section 3 of main.tex

\title{Reducing Acceptance Marks in Emerson-Lei Automata by QBF Solving}
\author{Tereza Schwarzov\'{a}}{%Faculty of Informatics,
  Masaryk University, Brno, Czech Republic}{xschwar3@mail.muni.cz}{}{}
\author{Jan Strej\v{c}ek}{%Faculty of Informatics,
  Masaryk University, Brno, Czech Republic}{strejcek@fi.muni.cz}{https://orcid.org/0000-0001-5873-403X}{}
\author{Juraj Major}{%Faculty of Informatics,
  Masaryk University, Brno, Czech Republic}{major@fi.muni.cz}{}{}
\authorrunning{T. Schwarzov\'{a}, J. Strej\v{c}ek, and J. Major}
\Copyright{Tereza Schwarzov\'{a}, Jan Strej\v{c}ek, and Juraj Major}
\ccsdesc[500]{Theory of computation~Logic}
\ccsdesc[500]{Theory of computation~Automata over infinite objects}
\keywords{Emerson-Lei automata, TELA, automata reduction, QBF, telatko}
%\funding{This work has been supported by the Czech Science Foundation grant GA19-24397S.}

@@

TELA have attracted a lot of attention during the last %five
years~\cite{baier.19.atva,major.19.atva,muller.17.gandalf,renkin.20.atva}.
Their popularity comes probably from the fact that these automata can
often use fewer states than equivalent automata with simpler
acceptance conditions. Further, algorithms handling TELA can
automatically handle all automata with traditional acceptance
conditions. TELA can be obtained for example by translating formulas
of \emph{linear temporal logic (LTL)}~\cite{pnueli.77.focs} with tools
\texttt{ltl2dela} (known as Delag)~\cite{muller.17.gandalf} or
\texttt{ltl2dgra} (known as Rabinizer~4)~\cite{kretinsky.18.cav} of
the Owl library, \texttt{ltl2tgba} of the Spot
library~\cite{duret.16.atva2}, or
\texttt{ltl3tela}~\cite{major.19.atva}. There are also algorithms
processing these automata, for example the emptiness
check~\cite{baier.19.atva} or translation of TELA to parity
automata~\cite{renkin.20.atva,casares.22.tacas}.

@@

Recall that we aim to reduce the number of acceptance marks in a given
TELA $\mA =(Q,M,\Sigma,\delta,q_I,\varphi)$ without altering its
structure and language. In other words, we look for
\begin{itemize}
\item a set $\oM$ of acceptance marks satisfying $|\oM|<|M|$,
\item an acceptance formula $\psi$ over $\oM$, and 
\item a function
  $\rem:\delta\rightarrow Q \times \Sigma \times 2^{\oM}\times Q$
  assigning new marks to transitions (i.e., for each
  $t=(p,a,M',q)\in\delta$, we assume that $\rem(t)=(p,a,\oM',q)$ for
  some $\oM'\subseteq\oM$)
  %
  such that the automaton $\mB=(Q,\oM,\Sigma,\rem(\delta),q_I,\psi)$
  is equivalent to $\mA$.
\end{itemize}
To ensure that the modified automaton is equivalent to the original one,
we actually look for $\psi$ and $\rem$ such that each run
$\pi=t_0t_1t_2\ldots$ of $\mA$ is accepting if and only if the run
$\rem(t_0)\rem(t_1)\rem(t_2)\ldots$ of $\mB$ is accepting.

We start with several simple observations. Acceptance of a run $\pi$
is fully determined by the set $\minf(\pi)$.
% of acceptance marks that
% appear infinitely often on $\pi$, which is in turn fully determined by
% the set $\minf(\pi)$ of transitions appearing infinitely often in
% $\pi$.
Further, each set $\minf(\pi)$ for a run $\pi$ is a cycle and vice
versa. Hence, our reduction method is looking for a new acceptance
formula $\psi$ and function $\rem$ such that for each cycle
$T\subseteq\delta$, it holds that $T$ satisfies $\varphi$ if and only
if $\rem(T)$ satisfies $\psi$. This can be roughly denoted by the
formula
\[
  \forall T\subseteq\delta~.~\mcycle(T)\implies\big(\msat_\varphi(T) ~\iff~ \msat_{\psi}(\rem(T))\big).
\]
In fact, this corresponds to the shape of the QBF we will
construct. As we are looking for $\psi$ and $\rem$ such that the
formula holds, the subformula $\msat_{\psi}(\rem(T))$ contains many
free variables representing possible instances of $\psi$ and
$\rem$. If the formula is satisfiable, then each of its models encodes
a desired instance of $\psi$ and $\rem$. In the following, we assume
that we are looking for a new acceptance formula $\psi$ in DNF. The
choice of DNF is not fundamental, but inherited from our previous
attempt to reduce acceptance formulas. The presented method can be
easily adapted to look for $\psi$ in \emph{conjunctive normal form
  (CNF)} or in a different shape.

>>>>

\title{Reducing Acceptance Marks in Emerson-Lei Automata by QBF Solving}
\author{Tereza Schwarzov\'{a}}{%Faculty of Informatics,
  Masaryk University, Brno, Czech Republic}{xschwar3@mail.muni.cz}{}{}
\author{Jan Strej\v{c}ek}{%Faculty of Informatics,
  Masaryk University, Brno, Czech Republic}{strejcek@fi.muni.cz}{https://orcid.org/0000-0001-5873-403X}{}
\author{Juraj Major}{%Faculty of Informatics,
  Masaryk University, Brno, Czech Republic}{major@fi.muni.cz}{}{}
\authorrunning{T. Schwarzov\'{a}, J. Strej\v{c}ek, and J. Major}
\Copyright{Tereza Schwarzov\'{a}, Jan Strej\v{c}ek, and Juraj Major}
\ccsdesc[500]{Theory of computation~Logic}
\ccsdesc[500]{Theory of computation~Automata over infinite objects}
\keywords{Emerson-Lei automata, TELA, automata reduction, QBF, telatko}
\funding{T.~Schwarzová received funding from
the European Union’s Horizon Europe program under the Grant Agreement
No.\ 101087529.}

@@

TELA have attracted a lot of attention during the last few
years~\cite{baier.19.atva,major.19.atva,muller.17.gandalf,renkin.20.atva}.
Their popularity comes probably from the fact that these automata can
often use fewer states than equivalent automata with simpler
acceptance conditions. Further, algorithms handling TELA can
automatically handle all automata with traditional acceptance
conditions. TELA can be obtained for example by translating formulas
of \emph{linear temporal logic (LTL)}~\cite{pnueli.77.focs} with tools
\texttt{ltl2dela} (known as Delag)~\cite{muller.17.gandalf} or
\texttt{ltl2dgra} (known as Rabinizer~4)~\cite{kretinsky.18.cav} of
the Owl library, \texttt{ltl2tgba} of the Spot
library~\cite{duret.16.atva2}, or
\texttt{ltl3tela}~\cite{major.19.atva}. There are also algorithms
processing these automata, for example the emptiness
check~\cite{baier.19.atva} or translation of TELA to parity
automata~\cite{renkin.20.atva,casares.22.tacas}.

@@

Recall that we aim to reduce the number of acceptance marks in a given
TELA $\mA =(Q,M,\Sigma,\delta,q_I,\varphi)$ without altering its
structure and language. In other words, we look for
\begin{itemize}
\item a set $\oM$ of acceptance marks satisfying $|\oM|<|M|$,
\item an acceptance formula $\psi$ over $\oM$, and 
\item a function
  $\rem:\delta\rightarrow Q \times \Sigma \times 2^{\oM}\times Q$
  assigning new marks to transitions (i.e., for each
  $t=(p,a,M',q)\in\delta$, we assume that $\rem(t)=(p,a,\oM',q)$ for
  some $\oM'\subseteq\oM$)
  %
  such that the automaton $\mB=(Q,\oM,\Sigma,\rem(\delta),q_I,\psi)$
  is equivalent to $\mA$.
\end{itemize}
% To simplify the task, 
% To ensure that the modified automaton is equivalent to the original one,
We will actually look for $\psi$ and $\rem$ such that each run
$\pi=t_0t_1t_2\ldots$ of $\mA$ is accepting if and only if the run
$\rem(t_0)\rem(t_1)\rem(t_2)\ldots$ of $\mB$ is accepting. This
requirement clearly guarantees the equiavalence of $\mA$ and $\mB$. We
note that preserving the acceptance of individual runs is not a
necessary condition for language equivalence of $\mA$ and $\mB$ and
there exist automata where relaxing this requirement can lead to
a bigger reduction of acceptance marks (see
Figure~\ref{fig:counter}). However, looking for $\psi$ and $\rem$
preserving the acceptance of individual runs makes the problem easier
as we can, for example, ignore the labelling of transitions by the
elements of $\Sigma$.
\begin{figure}[t]
  \centering
  \begin{tikzpicture}[smallautomaton]
    \begin{scope}[initial angle=180,xshift=-3.5cm]
      \node[state,initial] (p) at (0,0) {};
      \node[state] (q) at (2,0) {};
      \path[->,auto]
      (p) edge[out=120,in=60,looseness=10] pic {acc={0}{0}} pic {l=$a$} (p)
      (p) edge[out=-60,in=-120,looseness=10] pic {acc={1}{1}} pic {l=$b$} (p)
      (p) edge[out=30,in=150] pic {l=$a$} (q)
      (q) edge[out=-150,in=-30] pic {l=$b$} (p);

      \node[acclabel] at (1,-1.7)
      {$\Inf\tacc{0}{0}\wedge\Inf\tacc{1}{1}$};
    \end{scope}
    \begin{scope}[initial angle=180,xshift=3.5cm]
      \node[state,initial] (p) at (0,0) {};
      \node[state] (q) at (2,0) {};
      \path[->,auto]
      (p) edge[out=120,in=60,looseness=10] pic {l=$a$} (p)
      (p) edge[out=-60,in=-120,looseness=10] pic {l=$b$} (p)
      (p) edge[out=30,in=150] pic {acc={0}{0}} pic {l=$a$} (q)
      (q) edge[out=-150,in=-30] pic {l=$b$} (p);

      \node[acclabel] at (1,-1.7)
      {$\Inf\tacc{0}{0}$};
    \end{scope}
  \end{tikzpicture}
  \smallskip

  \caption{Both automata accept the same language of words
    $w\in\{a,b\}^\omega$ that contain infinitely many occurences of
    both $a$ and $b$. Each accepting run of the left automaton has to
    contain infinitely many occurrences of both transitions looping on
    the initial state. Hence, there does not exist any automaton with
    the same accepting runs as the left automaton and less than two
    accetpance marks.}
  \label{fig:counter}
\end{figure}

\fixme{TADY TO HONZA JESTE UPRAVUJE}
We start with several simple observations. Acceptance of a run $\pi$
is fully determined by the set $\minf(\pi)$.
% of acceptance marks that
% appear infinitely often on $\pi$, which is in turn fully determined by
% the set $\minf(\pi)$ of transitions appearing infinitely often in
% $\pi$.
\begin{lemma}
  A set $T\subseteq\delta$ is a cycle if and only if
  there is a run $\pi$ such that $T=\minf(\pi)$.
\end{lemma}

As $\minf(\pi)$ is the set of transitions that appear infinitely many
times in $\pi$, there has to be a suffix $\pi'$ of $\pi$ containing
only transitions of $\minf(\pi)$. Let $p$ be the first state of
$\pi'$. As each transition of $\pi'$ appears infinitely many times in
$\pi$ and thus also in $\pi'$, there has to be a finite prefix $\rho$
of $\pi'$ such that $\rho$ is a path from $p$ to $p$ that contains all
transitions of $\minf(\pi)$. In other words, the set $\minf(\pi)$ of a
run $\pi$ is always a cycle. Moreover, as we assume that 
%\footnote{We can also prove that for each cycle $T$ there exists a run $\pi$ such that $T=\minf(\pi)$.}

\fixme{ODTUD DAL JE TO PUVODNI, ale taky se to trochu bude menit!}

Further, each set $\minf(\pi)$ for a run $\pi$ is a cycle and vice
versa. Hence, our reduction method is looking for a new acceptance
formula $\psi$ and function $\rem$ such that for each cycle
$T\subseteq\delta$, it holds that $T$ satisfies $\varphi$ if and only
if $\rem(T)$ satisfies $\psi$. This can be roughly denoted by the
formula
\[
  \forall T\subseteq\delta~.~\mcycle(T)\implies\big(\msat_\varphi(T) ~\iff~ \msat_{\psi}(\rem(T))\big).
\]
In fact, this corresponds to the shape of the QBF we will
construct. As we are looking for $\psi$ and $\rem$ such that the
formula holds, the subformula $\msat_{\psi}(\rem(T))$ contains many
free variables representing possible instances of $\psi$ and
$\rem$. If the formula is satisfiable, then each of its models encodes
a desired instance of $\psi$ and $\rem$. In the following, we assume
that we are looking for a new acceptance formula $\psi$ in DNF. The
choice of DNF is not fundamental, but inherited from our previous
attempt to reduce acceptance formulas. The presented method can be
easily adapted to look for $\psi$ in \emph{conjunctive normal form
  (CNF)} or in a different shape.
